\TitlePt

### 2016-03-30 paper-fatiando.tex
All instances of Fatiando a Terra in italics

This chapter was published in the Proceedings of the 12th Python in Science
Conference in 2013
(\url{http://conference.scipy.org/proceedings/scipy2013/uieda.html}).
It describes the open-source software library Fatiando a Terra and the state of
the project at the time of the conference.
Fatiando a Terra was developed by me as part of my PhD thesis work.
The library has grown since 2013 and now contains more features than described
here.
For example, the algorithm for gravitational forward modeling with tesseroids
presented in Chapter~\ref{chap:tesseroids} is also implemented in Fatiando a
Terra.
Furthermore, the gravity inversion method presented in Chapter~\ref{chap:moho}
is implemented using functions and classes from Fatiando a Terra and will be
included in a future release of the library.
The official website (\url{http://www.fatiando.org}) contains more up-to-date
information about the project.

@@

Geophysics is the science of using physical observations of the Earth to infer
its inner structure. Generally, this is done with a variety of numerical
modeling techniques and inverse problems. The development of new algorithms
usually involves copy and pasting of code, which leads to errors and poor code
reuse. Fatiando a Terra is a Python library that aims to automate common tasks
and unify the modeling pipeline inside of the Python language. This allows
users to replace the traditional shell scripting with more versatile and
powerful Python scripting. The library can also be used as an API (Application
Programming Interface) for developing stand-alone programs.  Algorithms
implemented in Fatiando a Terra can be combined to build upon existing
functionality. This flexibility facilitates prototyping of new algorithms and
quickly building interactive teaching exercises. In the future, we plan to
continuously implement sample problems to help teach geophysics as well as
classic and state-of-the-art algorithms.

@@

Fatiando a Terra (\url{http://www.fatiando.org}, see
Figure~\ref{fig:p1-fatiando-site}) aims at providing such an API
for geophysical modeling. Functions in the \texttt{fatiando} package use
compatible data and mesh formats so that the output of one modeling function
can be used as input for another. Furthermore, routines can be combined and
reused to create new modeling algorithms.  Fatiando a Terra also automates
common tasks such as griding, map plotting with Matplotlib \citep[][
\url{http://matplotlib.org}]{hunter2007}, and 3D plotting with Mayavi \citep[][
\url{http://code.enthought.com/projects/mayavi}]{ramachandran2011}.  Version
0.1\footnote{
    This was the current version in 2013 when this chapter was published in the
    conference proceedings. As of April 2016, the latest version is 0.3.}
of Fatiando a Terra is focused on gravity and magnetic methods because this is
the main focus of the developers.  However, simple ``toy'' problems for
seismology and geothermics are available and can be useful for teaching
geophysics.

@@

The modules and packages of Fatiando a Terra are bundled into the
\texttt{fatiando} package. Each type of geophysical method has its own
package. As of version 0.1, the available modules and packages are:

@@

Fatiando a Terra handles map data as 1D Numpy arrays, typically x-, y-,
z-coordinates and an extra array with the corresponding data. However,
Matplotlib functions, like \texttt{contourf} and \texttt{pcolor},
require data to be passed as 2D arrays. Moreover, geophysical data sets
are often irregularly sampled and require griding before they can be
plotted. Thus, griding and array reshaping are ideal targets for
automation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/paper-fatiando/gridding_plotting_contourf}
    \caption{
    Example of 1) generating a random scatter of points (black dots),
    2) using that to make synthetic data, and
    3) automatically gridding and plotting the data using a
    Fatiando a Terra wrapper for the Matplotlib ``contourf``
    function.
    }
    \label{fig:p1-contourf}
\end{figure}

@@

The representation of 2D and 3D geometric elements is handled by the
classes in the \texttt{fatiando.mesher} module. Geometric elements in
Fatiando a Terra can be assigned physical property values, like density,
magnetization, seismic wave velocity, impedance, etc. This is done
through a \texttt{props} dictionary whose keys are the name of the
physical property and values are the corresponding values in SI units:

@@

The Fatiando a Terra packages have separate modules for forward modeling
and inversion algorithms. The forward modeling functions usually take as
arguments geometric elements from \texttt{fatiando.mesher} with assigned
physical properties and return the synthetic data. For example, the
module \texttt{fatiando.gravmag.tesseroid} is a Python implementation of
the program Tesseroids (\url{http://leouieda.github.io/tesseroids}) and
calculates the gravitational fields of tesseroids (e.g., spherical
prisms). The following example shows how to calculate the gravity
anomaly of the tesseroid model generated in the previous section
(Figure~\ref{fig:p1-tesseroidgrav}):

@@

Geophysics uses anomalies in the gravitational and magnetic fields
generated by density and magnetization contrasts within the Earth to
investigate the inner Earth structure. The Fatiando a Terra 0.1 release
has been focused on gravity and magnetic methods. Therefore, the
\texttt{fatiando.gravmag} package contains more advanced and
state-of-the-art algorithms than the other packages.

@@

Also implemented in Fatiando a Terra are some recent developments in
gravity and magnetic inversion methods. The method of ``planting
anomalous densities'' by \citet{uieda2012} is implemented in the
\texttt{fatiando.gravmag.harvester} module. In contrast to imaging
methods, this is an inversion method, i.e., it estimates a physical
property distribution (density in the case of gravity data) that fits
the observed data. This particular method requires the user to specify a
``seed'' (Figure~\ref{fig:p1-seed}) around which the estimated density
distribution grows (Figure~\ref{fig:p1-harvester}):

@@

The Fatiando a Terra package provides an API to develop modeling
algorithms for a variety of geophysical methods. The current version
(0.1)\footnote{
    As of April 2016, the latest version is 0.3.}
has a few state-of-the-art gravity and magnetic modeling and
inversion algorithms. There are also toy problems in gravity, seismics
and seismology that are useful for teaching basic concepts of
geophysics, modeling, and inverse problems.

Fatiando a Terra enables quick prototyping of new algorithms because of
the collection of fast forward modeling routines and the simple syntax
and high level of the Python language. After prototyping, the
performance bottlenecks of these algorithms can be easily diagnosed
using the advanced profiling tools available in the Python language.
Optimization of only small components of code can be done without loss
of flexibility using the Cython language \citep{behnel2011}.

The biggest challenge that Fatiando a Terra faces in the near future is
the development of a user and, consequently, a developer community. This
is a key part for the survival of any open-source project.

>>>>

This chapter was published in the Proceedings of the 12th Python in Science
Conference in 2013
(\url{http://conference.scipy.org/proceedings/scipy2013/uieda.html}).  It
describes the open-source software library \textit{Fatiando a Terra} and the
state of the project at the time of the conference.  \textit{Fatiando a Terra}
was developed by me as part of my PhD thesis work.  The library has grown since
2013 and now contains more features than described here.  For example, the
algorithm for gravitational forward modeling with tesseroids presented in
Chapter~\ref{chap:tesseroids} is also implemented in \textit{Fatiando a Terra}.
Furthermore, the gravity inversion method presented in Chapter~\ref{chap:moho}
is implemented using functions and classes from \textit{Fatiando a Terra} and
will be included in a future release of the library.  The official website
(\url{http://www.fatiando.org}) contains more up-to-date information about the
project.

@@

Geophysics is the science of using physical observations of the Earth to infer
its inner structure. Generally, this is done with a variety of numerical
modeling techniques and inverse problems. The development of new algorithms
usually involves copy and pasting of code, which leads to errors and poor code
reuse. \textit{Fatiando a Terra} is a Python library that aims to automate common tasks
and unify the modeling pipeline inside of the Python language. This allows
users to replace the traditional shell scripting with more versatile and
powerful Python scripting. The library can also be used as an API (Application
Programming Interface) for developing stand-alone programs.  Algorithms
implemented in \textit{Fatiando a Terra} can be combined to build upon existing
functionality. This flexibility facilitates prototyping of new algorithms and
quickly building interactive teaching exercises. In the future, we plan to
continuously implement sample problems to help teach geophysics as well as
classic and state-of-the-art algorithms.

@@

\textit{Fatiando a Terra} (\url{http://www.fatiando.org}, see
Figure~\ref{fig:p1-fatiando-site}) aims at providing such an API
for geophysical modeling. Functions in the \texttt{fatiando} package use
compatible data and mesh formats so that the output of one modeling function
can be used as input for another. Furthermore, routines can be combined and
reused to create new modeling algorithms.  \textit{Fatiando a Terra} also automates
common tasks such as griding, map plotting with Matplotlib \citep[][
\url{http://matplotlib.org}]{hunter2007}, and 3D plotting with Mayavi \citep[][
\url{http://code.enthought.com/projects/mayavi}]{ramachandran2011}.  Version
0.1\footnote{
    This was the current version in 2013 when this chapter was published in the
    conference proceedings. As of April 2016, the latest version is 0.3.}
of \textit{Fatiando a Terra} is focused on gravity and magnetic methods because this is
the main focus of the developers.  However, simple ``toy'' problems for
seismology and geothermics are available and can be useful for teaching
geophysics.

@@

The modules and packages of \textit{Fatiando a Terra} are bundled into the
\texttt{fatiando} package. Each type of geophysical method has its own
package. As of version 0.1, the available modules and packages are:

@@

\textit{Fatiando a Terra} handles map data as 1D Numpy arrays, typically x-, y-,
z-coordinates and an extra array with the corresponding data. However,
Matplotlib functions, like \texttt{contourf} and \texttt{pcolor},
require data to be passed as 2D arrays. Moreover, geophysical data sets
are often irregularly sampled and require griding before they can be
plotted. Thus, griding and array reshaping are ideal targets for
automation.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{figures/paper-fatiando/gridding_plotting_contourf}
    \caption{
    Example of 1) generating a random scatter of points (black dots),
    2) using that to make synthetic data, and
    3) automatically gridding and plotting the data using a
    \textit{Fatiando a Terra} wrapper for the Matplotlib ``contourf``
    function.
    }
    \label{fig:p1-contourf}
\end{figure}

@@

The representation of 2D and 3D geometric elements is handled by the
classes in the \texttt{fatiando.mesher} module. Geometric elements in
\textit{Fatiando a Terra} can be assigned physical property values, like density,
magnetization, seismic wave velocity, impedance, etc. This is done
through a \texttt{props} dictionary whose keys are the name of the
physical property and values are the corresponding values in SI units:

@@

The \textit{Fatiando a Terra} packages have separate modules for forward modeling
and inversion algorithms. The forward modeling functions usually take as
arguments geometric elements from \texttt{fatiando.mesher} with assigned
physical properties and return the synthetic data. For example, the
module \texttt{fatiando.gravmag.tesseroid} is a Python implementation of
the program Tesseroids (\url{http://leouieda.github.io/tesseroids}) and
calculates the gravitational fields of tesseroids (e.g., spherical
prisms). The following example shows how to calculate the gravity
anomaly of the tesseroid model generated in the previous section
(Figure~\ref{fig:p1-tesseroidgrav}):

@@

Geophysics uses anomalies in the gravitational and magnetic fields
generated by density and magnetization contrasts within the Earth to
investigate the inner Earth structure. The \textit{Fatiando a Terra} 0.1 release
has been focused on gravity and magnetic methods. Therefore, the
\texttt{fatiando.gravmag} package contains more advanced and
state-of-the-art algorithms than the other packages.

@@

Also implemented in \textit{Fatiando a Terra} are some recent developments in
gravity and magnetic inversion methods. The method of ``planting
anomalous densities'' by \citet{uieda2012} is implemented in the
\texttt{fatiando.gravmag.harvester} module. In contrast to imaging
methods, this is an inversion method, i.e., it estimates a physical
property distribution (density in the case of gravity data) that fits
the observed data. This particular method requires the user to specify a
``seed'' (Figure~\ref{fig:p1-seed}) around which the estimated density
distribution grows (Figure~\ref{fig:p1-harvester}):

@@

The \textit{Fatiando a Terra} package provides an API to develop modeling
algorithms for a variety of geophysical methods. The current version
(0.1)\footnote{
    As of April 2016, the latest version is 0.3.}
has a few state-of-the-art gravity and magnetic modeling and
inversion algorithms. There are also toy problems in gravity, seismics
and seismology that are useful for teaching basic concepts of
geophysics, modeling, and inverse problems.

\textit{Fatiando a Terra} enables quick prototyping of new algorithms because of
the collection of fast forward modeling routines and the simple syntax
and high level of the Python language. After prototyping, the
performance bottlenecks of these algorithms can be easily diagnosed
using the advanced profiling tools available in the Python language.
Optimization of only small components of code can be done without loss
of flexibility using the Cython language \citep{behnel2011}.

The biggest challenge that \textit{Fatiando a Terra} faces in the near future is
the development of a user and, consequently, a developer community. This
is a key part for the survival of any open-source project.

### 2016-03-30 paper-moho.tex
Add online repository section to Moho paper as well
All instances of Fatiando a Terra in italics

The inversion method proposed here is implemented in the Python programming
language.
The software is freely available
under the terms of the BSD 3-clause open-source software license.
Our implementation relies on the open-source libraries
scipy and numpy \citep[][ \url{http://scipy.org}]{jones2001}
for array-based computations,
matplotlib \citep[][ \url{http://matplotlib.org}]{hunter2007}
and seaborn
\citep[][ \url{http://stanford.edu/~mwaskom/software/seaborn}]{waskom2015}
for plots and maps,
and Fatiando a Terra \citep[][ \url{http://www.fatiando.org}]{uieda2013a}
for geophysics specific tasks,
particularly for forward modeling using tesseroids.
We use the scipy.sparse package for sparse matrix arithmetic and linear
algebra.
The sparse linear system in Eq.~\ref{eq:moho-gaussnewton-regul}
is solved using the conjugate gradient method implemented in scipy.sparse.

@@

>>>>

The inversion method proposed here is implemented in the Python programming
language.
The software is freely available
under the terms of the BSD 3-clause open-source software license.
Our implementation relies on the open-source libraries
scipy and numpy \citep[][ \url{http://scipy.org}]{jones2001}
for array-based computations,
matplotlib \citep[][ \url{http://matplotlib.org}]{hunter2007}
and seaborn
\citep[][ \url{http://stanford.edu/~mwaskom/software/seaborn}]{waskom2015}
for plots and maps,
and \textit{Fatiando a Terra} \citep[][ \url{http://www.fatiando.org}]{uieda2013a}
for geophysics specific tasks,
particularly for forward modeling using tesseroids.
We use the scipy.sparse package for sparse matrix arithmetic and linear
algebra.
The sparse linear system in Eq.~\ref{eq:moho-gaussnewton-regul}
is solved using the conjugate gradient method implemented in scipy.sparse.

@@

\section{Online repository}

The Moho depth model for South America estimated here can be downloaded from
the online repository
\url{https://github.com/pinga-lab/paper-moho-inversion-tesseroids}.
The repository also contains all data and source code used to produce the
results and figures presented here.
